---
title: "Home work 1"
output:
  html_document: default
  pdf_document: default
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache=TRUE)
library(tidyverse)
```

# **Homework 1**

**Names**: Abdullah Faqih AlMubarok,August Nygaard Bodilsen, Natalie
Christiansen, Maria Madrazo I Montoya

**Group**: 21

## **Question 1**

Install the package babynames and look at the data babynames:

```{r, eval=FALSE}
install.packages("babynames")
```

```{r q1.part1}
library(babynames)
head(babynames)
```

a)  List the top 5 female baby names starting with P, regardless of
    year, as a table.

```{r}
result1<-babynames %>%
  filter(sex=="F" & str_detect(name, "^P"))%>%
  group_by(name)%>%
  summarise(total = sum(n))%>%
  arrange(desc(total))%>%
  head(5)

result.table <- result1%>%
  as.matrix()%>%
  as.table()
row.names(result.table)<-c(1:5)
result.table
```

b)  Using the results from a, plot their occurrences as a function of
    year using a line plot. Comment on your results. If you get strange
    results, explain them and/or improve the plot.

```{r}
  
result2<- result1$name
top5 <- babynames %>%
         filter(name==result2 & sex=="F" )
ggplot(top5, aes(x = year, y = n)) +
  geom_line(aes(col = name)) +
  xlab("Year") +
  ylab("Frequence (number of babies)") + 
  ggtitle("Top five female baby names") +
  theme_bw()

```

The plot shows that the most common female name starting with P was
Patricia, between 1930 and 1950. In second place, it was Pamela around
1940 and 1970. The rest of the names had similar ocurrences. After 1950,
the level of ocurrence for these names gradually dropped to very low
levels. The absence of values before 1910 could be due to lack of
registration or low popularity of the names.

The graphs in the plot overlap, this could be solved by separating the
graphs based on name.

```{r}
ggplot(top5, aes(x = year, y = n)) +
  geom_line() +
  xlab("Year") +
  ylab("Frequence (number of babies)") + 
  ggtitle("Top five female baby names") +
  facet_wrap(~name, scales="free")+
  theme_bw()
```

## **Question 2**

In the same dataset, is the name Arwen significantly more (or less)
common in 2004 vs 1990? Is the change significant? What is the likely
cause? Do not use hard-coding.

We use fischers exact test to compare if the count of girls named Arwen
in 2004 significantly differs from the count in 1990. In other words, we
will test if the relative proportion of babies named Arwen in 2004 is
different than in 1990. We use fishers exact test instead of the
chi-square test, due to the count of people named Arwen being very
low.'\

**H0**: The odds ratio is 1. There is no relationship between rows and
columns in our test matrix. \
**Halt**: The odds ratio is not 1.\
**alpha**= 0.05

```{r}
babynames %>%
  group_by(year) %>%
  mutate(other_name = sum(n)-n) %>%
  filter(name == "Arwen", year == 1990|year == 2004) %>%
  arrange(desc(year)) %>%
  select(year, n, other_name) -> edited_babynames

fisher_test_matrix = as.matrix(column_to_rownames(edited_babynames, "year"))
fisher_test_matrix
fisher.test(fisher_test_matrix)
```

Since we obtain a p-value \< 2.2e-16, which is \< alpha, we can reject
the null hypothesis. So, there is a significant difference (an increase)
in the number of babies named Arwen in 2004 compared to 1990. This could
be due to the release of the movies "Lord of The Rings", since one of
the characters had this name.

**Conclusion**: The P-value is \< 2.2e-16, so we can be very confident
that the odds ratio is not 1. Our estimated odds ratio is 17.17996,
indicating that there is a much larger chance of being named Arwen when
born in 1990 compared to being born in 2004. This could be due to the
release of the movies "Lord of The Rings", since one of the characters
had this name.

## **Question 3**

Produce the following plot starting from the flowers dataset. A
potentially useful function that you may not have seen: bind_rows():
merges two tibbles by rows so that the joint tibble becomes longer, not
wider

```{r}
iris_petal <- iris %>% select(starts_with("Petal"), Species) %>%
  mutate(Dimension = "Petal", width = Petal.Width, length = Petal.Length) %>%
  select(Species, length, width, Dimension)

iris_sepal <- iris %>% select(starts_with("Sepal"), Species) %>%
  mutate(Dimension = "Sepal", width = Sepal.Width, length = Sepal.Length) %>%
  select(Species, length, width, Dimension)

bind_rows(iris_petal, iris_sepal) %>% 
  ggplot(aes(y=width, x=length, col=Species)) + geom_point() +
  geom_smooth(method="lm", fullrange=TRUE) + facet_wrap(~Dimension, scales="free") + 
  theme_bw()+ggtitle("Question 3")
```

## **Question 4**

We are given a file with binding sites of a certain transcription
factor, made with the ChIP-seq technique (you will hear a lot more about
the technique later in the course) by a collaborator. In the homework
directory, there is a data file 'chip_mm5.txt' from the collaborator,
representing binding sites from a Chip-chip experiment, with a column
for chromosome, start, end, and score, where score is how 'good' the
binding is. Our collaborator has two hypotheses:

1: Binding scores are dependent on chromosome \
2: Binding site widths (end-start) are dependent on chromosome

Can you prove/disprove these two hypotheses statistically?

```{r}
chip<- read_tsv("chip_mm5.txt")
```

*Hypothesis 1: Binding scores are dependent on chromosome*

To answer the first hypothesis, we need to compare if any of the
chromosomes are different in means for binding scores.

**H0**= There are no differences in mean binding score between
chromosomes \
**Halt**= The mean binding scores differ between the chromosomes, \
**alpha**= 0.05

First, we inspect the distribution of scores for the individual
chromosomes:

```{r}
chip %>% ggplot(aes(x=score)) + geom_histogram(alpha=0.5) + facet_wrap(~chr) +
theme_bw()+
ggtitle("ChIP-seq scores by chromosome")
```

The test of choice will be one-way ANOVA test because we are working
with a continuous dependent variable (score) and one categorical
predictor (chromosome), as well as the assumption of normally
distributed scores.

```{r}
oneway.test(score ~ as.factor(chr), data=chip)
```

Binding score does not seem to be dependent on chromosome. With a
p-value of 0.4298 (above our threshold), we cannot reject the null
hypothesis.

*Hypothesis 2: Binding site widths (end-start) are dependent on
chromosome*\
To answer the second hypothesis, we will create a new column called
"width", which takes into account the start and end of the binding. Once
again, we will test if width follows a normal-like distribution on the
chromosomes.

```{r}
chip=mutate(chip, width=end-start)
ggplot(chip,aes(x=width))+
  geom_density()+theme_bw()+facet_wrap(~chr)
```

In this case, the binding site width does not follow a normal
distribution. Because we have more than three chromosome groups to
analyze, we will be using the non-parametric version: Kruskal Wallis.

**H0**= The mean ranks of binding site widths are equal between
chromosomes \
**Halt**= The mean ranks of binding site widths are not equal between
chromosomes, alpha = 0.05

```{r}
kruskal.test(width ~ as.factor(chr), data=chip)
```

We get a p-value of 0.003288 (below our threshold), so we reject the
null hypothesis. Therefore, binding site width does seem to be dependent
on chromosome.
